---
output:
  xaringan::moon_reader:
    self_contained: false
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

<style>
h1   {color: rgb(0, 128, 128);}
h2   {color: rgb(0, 128, 128);}
h3   {color: rgb(0, 128, 128);}
h4   {color: rgb(0, 128, 128);}

/* .remark-slide-content { */
/*   padding-top: 20px; */
/*   padding-left: 20px; */
/*   padding-right: 20px; */
/*   padding-bottom: 20px; */
/* } */
.remark-slide-content > h1:first-of-type {
  margin-top: 0px;
  margin-bottom: 0px;
}
.remark-slide-content > h2:first-of-type {
  margin-top: 0px;
  margin-bottom: 0px;
}

a:link {
    color:  #1f618d;
}

a:visited {
    color:  #1f618d;
}


.small { font-size: 70% }
</style>

# An introduction to R markdown for reproducible research


<br>

```
Stephen J Eglen                  Cambridge Computational Biology Institute
https://sje30.github.io          University of Cambridge
sje30@cam.ac.uk                  @StephenEglen
```


Slides: http://bit.ly/eglen2018-rmd (CC-BY license)


<br>

#### Acknowledgements

Paul Charlseworth, Ellese Cotterill, Catherine Cutts, Tom Edinburgh,
BBSRC, EPSRC, Wellcome Trust, Software Sustainability Institute.


<!-- 
(Last update `r date()`) 
 -->


```{r,include=FALSE}
require(printr)
require(knitr)
require(xtable)
pbr = function(text) {
  x = '<div style="font-size: 60%; position: absolute; right: 15%;   width: fit-content; height: fit-content; top:95%; z-index:10">'
  sprintf("%s%s</div>", x, text)
}
doi = function(doi) {
  ##text=sprintf("[%s](http://dx.doi.org/%s)", doi, doi)
  text=sprintf('<a href="http://dx.doi.org/%s">%s</a>', doi, doi)
  pbr(text)
}
im = function(f, width=NULL, height=NULL) {
  if (!is.null(width)) {
    line = sprintf('<img src="%s", width=%s>',
                   knitr::image_uri(f), width)
  } else {
    if (!is.null(height)) {
      line = sprintf('<img src="%s", height=%s>',
                     knitr::image_uri(f), height)
    } else {
      line = sprintf('<img src="%s">', knitr::image_uri(f))
    }
  }
  line
}
im1 = function(f,width=600) {
  sprintf('<center><img src="%s", width=%s></center>\n', knitr::image_uri(f), width)
}
imh = function(f,height=600) {
  sprintf('<center><img src="%s", height=%s></center>\n', knitr::image_uri(f), height)
}
```


<!-- --- -->

<!-- <br> -->
<!-- <br> -->

<!-- `r im1('~/papers/talks/talkfigs/2017/prestige1975_title.png', width=730)` -->

---

# The reproducibility crisis

Many key findings in publications are either not independently
verified, or fail verification when it is attempted (Baker, 2016).
`r doi("10.1038/533452a")`

<br>
Duke oncogenomics scandal.  Awesome detective work by Keith Baggerley
and Kevin Coombes.
<https://www.youtube.com/watch?v=7gYIs7uYbMo>


<br>
<br>

Disclaimer: do I mean "reproducibility" or "replicability"?
(Barba 2018) <https://arxiv.org/pdf/1802.03311.pdf>


---


# Literate programming

Donald Knuth invented the literate programming environment to combine
code and documentation into one file.

In our context, we interweave R code and markdoen 


# Inverse problems are hard

.pull-left[
<br>
```{r,include=TRUE,echo=FALSE}
grades <- data.frame("Mark"= c("70-100", "60-69", "50-59", "40-49", "0-39"),
grade=LETTERS[c(1,2,3,4,6)])
kable(grades, format="html")
```
]
.pull-right[
### Forward problem

I scored 68, what was my grade?

### Inverse problem

I got a B, what was my score?
]


---
# Moral or selfish approach?

<br>

`r imh("../../talkfigs/2016/Markowetz2015-selfish.png", 400)`

`r doi("10.1186/s13059-015-0850-7")`
---
# Selfish reasons to share

Why not align what is good for science with what is good for
scientists?


1. Funding mandates (REF + enforcement from Wellcome Trust)
2. Credit through data papers
3. Fixes data bugs / errors in analysis
4. Prevent data loss [(Vines et al 2014)](http://www.ncbi.nlm.nih.gov/pubmed/24361065). e.g. students have a habit of leaving...
5. Your future self is probably one of the main beneficiaries of
   sharing.
6. *Now* is a very good time to be an open scientist.
7. Leads to further collaborations
8. Reviewers can do more work...


---
# Simple example of reproducible research

Eglen SJ (2016) Bivariate spatial point patterns in the retina: a
reproducible review. Journal de la Société Française de Statistique
157:33–48.

`r imh("~/papers/talks/2017/2017-01-16-cam/figs/eg-eglen2015-rr.png", 400)`

See
<!-- [paper](http://journal-sfds.fr/index.php/J-SFdS/article/view/518/490) -->
[paper](https://doi.org/10.1101/029348)
or [code](http://github.com/sje30/eglen2015).  [Docker image](https://hub.docker.com/r/sje30/eglen2015/).



---
# New tools

1. **Docker** Can bundle entire open-source evironment for others to share.
2. Jupyter notebooks
3. binder = Docker + jupyter + cloud compute
4. Code ocean, and alternatives, being supported by some journal publishers,
   e.g. CUP.

`r imh('~/papers/talks/2016/2016-07-04-copenhagen/figs/2p-ram-github.png',
300)`

[https://github.com/sofroniewn/2pRAM-paper](https://github.com/sofroniewn/2pRAM-paper)

---
# Binder example for teaching

Two hour introduction to computational neuroscience (this Thursday if
anyone is interested...)

[Binder](https://mybinder.org/v2/gh/sje30/julia-python/master?filepath=Introduction%20to%20Computational%20Neuroscience.ipynb)
[Github](https://github.com/sje30/julia-python)

---
# Find a code buddy

* We ask our students to submit a .Rnw file rather than a pdf.  You get
a zero if I can't compile the pdf.
* So, ask someone else if they can run your code.
* Bioconductor team performs code review
* Journals gradually moving in this direction


---
## Reproducible figures

<!-- Never mind making a whole paper reproducible, how about just one figure? -->

```{r patchwork,chunk=FALSE,include=FALSE}
##devtools::install_github("thomasp85/patchwork")
read_chunk('~/langs/R/patchwork.R',labels='p1')
```

```{r p1,message=FALSE,chunk=FALSE,fig.width=8,fig.height=5}
```

---
## Won't this be slow?

Q: "If R is always re-running the analysis every time I edit my paper,
won't this be slow?"

A: Yes.  In which case, use the **cache**.


---
## Rstudio features for working with Rmd files

---
# Summary

* Find the selfish reasons to make your research reproducible.
* Adopt good practices to help you on your way.
* Writing code in groups can be very motivating.
* Use new tech if you want, but old tech works too.

---
# Challenges

* Long computation times (CODE CHECK).
* licensed software complicates everything.
* Can journals handle reproducible documents?
* When is the best time to think reproducibly?  
   - Too early (explore  first) 
   - Too late (paper now out)?
* Technical challenges << Societal challenges
