---
output:
  xaringan::moon_reader:
    self_contained: false
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

<style>
h1   {color: rgb(0, 128, 128);}
h2   {color: rgb(0, 128, 128);}
h3   {color: rgb(0, 128, 128);}
h4   {color: rgb(0, 128, 128);}

/* .remark-slide-content { */
/*   padding-top: 20px; */
/*   padding-left: 20px; */
/*   padding-right: 20px; */
/*   padding-bottom: 20px; */
/* } */
.remark-slide-content > h1:first-of-type {
  margin-top: 0px;
  margin-bottom: 0px;
}
.remark-slide-content > h2:first-of-type {
  margin-top: 0px;
  margin-bottom: 0px;
}

a:link {
    color:  #1f618d;
}

a:visited {
    color:  #1f618d;
}


.small { font-size: 70% }
</style>

# An introduction to R markdown for reproducible research


<br>

```
Stephen J Eglen                  Cambridge Computational Biology Institute
https://sje30.github.io          University of Cambridge
sje30@cam.ac.uk                  @StephenEglen
```


Slides: http://bit.ly/eglen2018-rmd (CC-BY license)


<br>

#### Acknowledgements

Laurent Gatto, Mike Smith.

(Last update `r date()`) 



```{r,include=FALSE}
require(printr)
require(knitr)
require(xtable)
pbr = function(text) {
  x = '<div style="font-size: 60%; position: absolute; right: 15%;   width: fit-content; height: fit-content; top:95%; z-index:10">'
  sprintf("%s%s</div>", x, text)
}
doi = function(doi) {
  ##text=sprintf("[%s](http://dx.doi.org/%s)", doi, doi)
  text=sprintf('<a href="http://dx.doi.org/%s">%s</a>', doi, doi)
  pbr(text)
}
im = function(f, width=NULL, height=NULL) {
  if (!is.null(width)) {
    line = sprintf('<img src="%s", width=%s>',
                   knitr::image_uri(f), width)
  } else {
    if (!is.null(height)) {
      line = sprintf('<img src="%s", height=%s>',
                     knitr::image_uri(f), height)
    } else {
      line = sprintf('<img src="%s">', knitr::image_uri(f))
    }
  }
  line
}
im1 = function(f,width=600) {
  sprintf('<center><img src="%s", width=%s></center>\n', knitr::image_uri(f), width)
}
imh = function(f,height=600) {
  sprintf('<center><img src="%s", height=%s></center>\n', knitr::image_uri(f), height)
}
```

---
## Literate programming

Donald Knuth invented the literate programming environment to combine
code and documentation into one file.

In our context, we interweave R code and markdown.  R code is executed
and results (text, figures, tables) fed back into markdown.  Markdown
then translated into html (or pdf or Word or slides...).

`r imh("rmd-to-md.png", 400)`




---
## Simple example of reproducible research

Eglen SJ (2016) Bivariate spatial point patterns in the retina: a
reproducible review. Journal de la Société Française de Statistique
157:33–48.

`r imh("~/papers/talks/2017/2017-01-16-cam/figs/eg-eglen2015-rr.png", 400)`

See
<!-- [paper](http://journal-sfds.fr/index.php/J-SFdS/article/view/518/490) -->
[paper](https://doi.org/10.1101/029348)
or [code](http://github.com/sje30/eglen2015).  [Docker image](https://hub.docker.com/r/sje30/eglen2015/).

See other examples at: <https://rmarkdown.rstudio.com/gallery.html>


---
## What is markdown?

A light-weight markup language for generating HTML.

```
### Example

Here is some *markup text* with **bold** and a
[link](http://www.rstudio.org).
Maths can be included $x^2 + y^2 = z^2$.
```

### Example

Here is some *markup text* with **bold** and a
[link](http://www.rstudio.org).
Maths can be included $x^2 + y^2 = z^2$.

---
## Embedding code: inline

Take a chunk like:

Simple snippets of code can be embedded like ​‘r sqrt(144)‘ is the
square root of 144. 

(NB: the backtick character above.)

which makes:

Simple snippets of code can be embedded like `r sqrt(144)` is the
square root of 144.

---
## Embedding code: chunks

Or a chunk of code

    `​``{r eval=TRUE}
    n = 10
    rnorm(n)
    `​``

which makes this output:

```{r eval=TRUE}
n = 10
rnorm(n)
```


---

## Figures

If your code generates a figure, it is saved into `figures/` folder
and included in output.  (See bookdown for captions and references).

    ‘‘‘{r eval=TRUE, echo=TRUE, fig.height=4}
    x = rnorm(300, mean=4, sd=2)
    hist(x)
    ‘‘‘


```{r eval=TRUE,echo=TRUE,fig.height=4}
x = rnorm(300, mean=4, sd=2)
hist(x)
```


---
## Reproducible figures

Never mind making a whole paper reproducible, how about just one figure?


```{r patchwork,chunk=FALSE,include=FALSE}
##devtools::install_github("thomasp85/patchwork")
read_chunk('~/langs/R/patchwork.R',labels='p1')
```

```{r p1,message=FALSE,chunk=FALSE,fig.width=8,fig.height=5}
```

---
## Won't this be slow?

Q: "If R is always re-running the analysis every time I edit my paper,
won't this be slow?"

A: Yes.  In which case, use the **cache**.


---
## Rstudio demonstration

Create a new document with a simple structure.  See header and options
for compiling document.

---
## Bookdown

Bookdown sits atop knitr and provides extra features for writing a
book (captioning figures and tables, labels to refer to sections).

<https://bookdown.org/>  <https://bookdown.org/yihui/bookdown/>

---
## Challenges

* Long computation times (CODE CHECK).
* licensed software complicates everything.
* Can journals handle reproducible documents?
* When is the best time to think reproducibly?  
   - Too early (explore  first) 
   - Too late (paper now out)?
* Technical challenges << Societal challenges


---
## Extended example

Let's work through the document to estimate pi.

You will need `estimatek.Rmd` and `refs.bib`

(References are in BiBTeX format; you can get them from google
scholar, or e.g. from <https://www.bioinformatics.org/texmed/> which
is pubmed with BiBTeX.)


---
## rticles

The `rticles` package allows you to make Rmd templates for a growing
number of journals, easing the adoption of formatting for journals.

<https://github.com/rstudio/rticles>


---
## Next steps

* Material from today at: <http://github.com/sje30/2018-12-07-rmd>
* Jupyter <www.jupyter.org> as alternative format for notebook
  computation.
* Two-way Rmd <-> docx conversion: <https://noamross.github.io/redoc/>
* officedown <https://github.com/davidgohel/officedown>
* pandoc toolkit <https://pandoc.org/> drives a lot of conversion from
  markdown into other formats (EPUB, Word, PDF, XML...)
  





